Fast Random Integer Generation in an Interval

### 2023-12-04 Paper/main.tex
small changes

\subsection{Unbiased}
While all prior approaches are very simplistic and always require only a handful of instructions, they also generate a bias which can vary depending on $n$.
But since we want to generate a \emph{uniform} random number in $[0,n)$, we somehow need to get rid of such bias by incorporating a fail-safe: rejection sampling.

\subsubsection{OpenBSD}\label{sec:2.2.1}
We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
The key idea is that this interval does not have to be on the right.
Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ to achieve uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
This yields \hyperref[alg:openbsd]{Algorithm 4} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.

@@

While the bias is usually very small and negligible in most use-cases, if we sample millions of times, it can easily become apparent and strongly distort the final dataset depending on the algorithm.
Hence, it is important to generate unbiased random integers at the cost of a little execution time.
Common software libraries seem to rely on algorithms that always require at least one division, if not two or more which is very inefficient on most processors.
While Bitmask has in theory zero division operations, the required \texttt{clz} operation can be as inefficient as the \texttt{div} instruction.
Hence, the algorithm introduced by Lemire is a perfect fit for modern x64 processors as for small $n$, we require no division (or equally bad) operation at all and also bound the maximum number of divisions to at most one.

For different technologies however, support for the computation of the full multiplication might be lacking such as in Graphics Processing Units (GPUs)~\cite{GPU} or DRAM Processing Units (DPUs) used for In-Memory-Processing in the \textsc{Upmem} architecture~\cite{Upmem}.
In the \textsc{Upmem-Pim} toolchain, for example, an \texttt{clz} operation seems to greatly outshine a \texttt{div} operation~\cite{UpmemRng}.

>>>>

\subsection{Unbiased}
While all prior approaches are quite simplistic and always require only a handful of instructions, they also generate a bias which can vary depending on $n$.
But since we want to generate a \emph{uniform} random number in $[0,n)$, we somehow need to get rid of such bias by incorporating a fail-safe: rejection sampling.

\subsubsection{OpenBSD}\label{sec:2.2.1}
We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
The key idea is that this interval does not have to be on the right.
Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
This yields \hyperref[alg:openbsd]{Algorithm 4} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.

@@

While the bias is usually insignificant and negligible in most use-cases, if we sample millions of times, it can easily become apparent and strongly distort the final dataset depending on the algorithm.
Hence, it is important to generate unbiased random integers at the cost of a little execution time.
Common software libraries seem to rely on algorithms that always require at least one division, if not two or more which is inefficient on most processors.
While Bitmask has in theory zero division operations, the required \texttt{clz} operation can be as inefficient as the \texttt{div} instruction.
Hence, the algorithm introduced by Lemire is a perfect fit for modern x64 processors as for small $n$, we require no division (or equally bad) operation at all and also bound the maximum number of divisions to at most one.

For different technologies however, support for the computation of the full multiplication might be lacking such as in Graphics Processing Units (GPUs)~\cite{GPU} or DRAM Processing Units (DPUs) used for In-Memory-Processing in the \textsc{Upmem} architecture~\cite{Upmem, UpmemSDK}.
In the \textsc{Upmem-Pim} toolchain, for example, an \texttt{clz} operation seems to greatly outshine a \texttt{div} operation~\cite{UpmemRng}.
